---
title: "TK_63: Expression of Gna12 and Gna13 genes in mouse pancreatic beta cells"
author: "Hernan Lorenzi"
affiliation: "The TriLab Bioinformatics Group, LBG-LCMB-LBC, NIDDK, NIH"
date: "8/4/2023"
output:
  pdf_document: default
  html_document:
    df_print: paged
    theme:
      bootswatch: flatly
---

```{=html}
<style type="text/css">

h1.title {
  font-size: 38px;
  text-align: center;
}
h4.author {
  font-size: 24px;
  text-align: center;
}
h4.affiliation {
  font-size: 24px;
  text-align: center;
}
h4.date {
  font-size: 24px;
  text-align: center;
}
</style>
```
**PI:** Jurgen Wess\
**Point of Contact:** Srinivas Pittala\
**Contact email:**
[Srinivas.Pittala\@nih.gov](mailto:Contact_name@nih.gov){.email}

**Summary:** to find the murine G-proteins
[*GNA12*](https://www.uniprot.org/uniprotkb/P27600/entry) and
[*GNA13*](https://www.uniprot.org/uniprotkb/P27601/entry) expression levels in
pancreatic beta cells.

Candidate projects with expression info for G12/G13 proteins in pancreatic beta
cells are
[here](https://gcc02.safelinks.protection.outlook.com/?url=https%3A%2F%2Fwww.informatics.jax.org%2Fgxd%2Fhtexp_index%2Fsummary%3Fstructure%3Dpancreatic%2Bislet%26theilerStage%3D0%26_theilerStage%3D1%26age%3DANY%26_age%3D1%26mutatedIn%3D%26strain%3D%26sex%3D%26method%3D%26text%3D%26textScope%3DTitle%26textScope%3DDescription%26arrayExpressID%3D&data=05%7C01%7Chernan.lorenzi%40nih.gov%7Ce6ad33a500ca4594ac1208db9520044a%7C14b77578977342d58507251ca2dc2b06%7C0%7C0%7C638267736961442889%7CUnknown%7CTWFpbGZsb3d8eyJWIjoiMC4wLjAwMDAiLCJQIjoiV2luMzIiLCJBTiI6Ik1haWwiLCJXVCI6Mn0%3D%7C3000%7C%7C%7C&sdata=6PdBA9J0oLytLJ4xWeaUumRiHfvxNJm1cEvGjbXXXAQ%3D&reserved=0).
From these studies the analysis was run on WT samples from the following two
datasets:
[GSE16048](https://www.ncbi.nlm.nih.gov/projects/geo/query/acc.cgi?acc=GSE16048)
and [GSE29674](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE29674).

For dataset
[GSE16048](https://www.ncbi.nlm.nih.gov/projects/geo/query/acc.cgi?acc=GSE16048)
islets were isolated and RNA prepared from 8-week-old (gender not specified).
All the mice were bred on a mixed genetic background (SV 129/C57BL/6)

For dataset
[GSE29674](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE29674) islets
were isolated and RNA prepared from five 5-week-old female mice (C57/129 genetic
background).

### R code

```{r include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning=FALSE, error=FALSE)
```

### Load libraries

```{r}
library(tidyverse)
```

### Load dataset GSE16048

```{r}
# Load ensemblid2affyid file
ensembl2affy <- read_delim(file = "./data/ensemblid2affyid.txt", delim = "\t", col_names = c("EnsemblID", "Affy_MOE430A_ID", "Affy_MOE430B_ID"), skip = 1)

# Only load WT samples
# VALUES represent Log2 RMA normalized expression signal
for (sample in c("GSM401525","GSM401526","GSM401527")){ 
  assign(x = sample, value = read_delim(file = paste0("https://www.ebi.ac.uk/biostudies/files/E-GEOD-16048/",sample,"_sample_table.txt"), delim = "\t"))
}

# Merge datasets
data.16048 <- tibble(Probe=GSM401525$ID_REF, GSM401525=GSM401525$VALUE, GSM401526=GSM401526$VALUE, GSM401527=GSM401527$VALUE)

# Add column with median expression value per probe
data.16048 <- data.16048 %>% 
        rowwise() %>% 
        mutate(median=median(c(GSM401525,GSM401526,GSM401527)))

# Add column with mean expression value per probe
data.16048 <- data.16048 %>% 
        rowwise() %>% 
        mutate(mean=mean(c(GSM401525,GSM401526,GSM401527)))

data.16048_gene <- inner_join(x=data.16048, y=ensembl2affy, by=join_by(Probe == Affy_MOE430A_ID))
```

### Load dataset GSE29674

```{r}
# Only load WT samples
# VALUES represent Log2 RMA normalized expression signal
for (sample in c("GSM735944","GSM735945","GSM735946","GSM735947","GSM735948")){ 
  assign(x = sample, value = read_delim(file = paste0("https://www.ebi.ac.uk/biostudies/files/E-GEOD-29674/",sample,"_sample_table.txt"), delim = "\t"))
}

# Merge datasets
data.29674 <- tibble(Probe=GSM735944$ID_REF, 
                     GSM735944=GSM735944$VALUE, 
                     GSM735945=GSM735945$VALUE,
                     GSM735946=GSM735946$VALUE,
                     GSM735947=GSM735947$VALUE,
                     GSM735948=GSM735948$VALUE,)

# Add column with median expression value per probe
data.29674 <- data.29674 %>% 
        rowwise() %>% 
        mutate(median=median(c(GSM735944,GSM735946,GSM735946,GSM735947,GSM735948)))

# Add column with mean expression value per probe
data.29674 <- data.29674 %>% 
        rowwise() %>% 
        mutate(mean=mean(c(GSM735944,GSM735946,GSM735946,GSM735947,GSM735948)))

data.29674_gene <- inner_join(x=data.29674, y=ensembl2affy, by=join_by(Probe == Affy_MOE430A_ID))
```

### Summarize mean expression by gene (dataset GSE16048)

```{r}
mean_exp_per_gene.GSE16048 <- data.16048_gene %>% group_by(EnsemblID) %>% summarise(Mean_exp=mean(mean)) %>% arrange(desc(Mean_exp)) %>% mutate(Dataset="GSE16048") %>% mutate(Percent_rank=percent_rank(Mean_exp))

gna12 <- 'ENSMUSG00000000149'
gna13 <- 'ENSMUSG00000020611'

# Print percent rank for gna12 and gna13
print(filter(mean_exp_per_gene.GSE16048, EnsemblID  %in%  c(gna12, gna13)))
```

### Summarize mean expression by gene (dataset GSE29674)

```{r}
mean_exp_per_gene.GSE29674 <- data.29674_gene %>% group_by(EnsemblID) %>% summarise(Mean_exp=mean(mean)) %>% arrange(desc(Mean_exp)) %>% mutate(Dataset="GSE29674") %>% mutate(Percent_rank=percent_rank(Mean_exp))

gna12 <- 'ENSMUSG00000000149'
gna13 <- 'ENSMUSG00000020611'

# Print percent rank for gna12 and gna13
print(filter(mean_exp_per_gene.GSE29674, EnsemblID  %in%  c(gna12, gna13)))
```

### Results

| **EnsemblID**      | **Gene name** | **Mean expression** | **Percentile rank** | **Study** |
|----------------|:--------------:|:--------------:|:--------------:|:--------------:|
| ENSMUSG00000020611 |     gna13     |      5.298699       |      0.4688176      | GSE29674  |
| ENSMUSG00000000149 |     gna12     |      3.490793       |      0.3242475      | GSE29674  |
| ENSMUSG00000020611 |     gna13     |      8.245162       |      0.6560356      | GSE16048  |
| ENSMUSG00000000149 |     gna12     |      6.405584       |      0.4173311      | GSE16048  |

: Pancreatic beta-cell gene expression[^1] and percentile rank[^2] of GNA12 and
GNA13 genes across WT samples from studies GSE29674 and GSE16048.

[^1]: Log2 RMA normalized expression signal

[^2]: Percentage of expression values in its frequency distribution that are
    less than a given gene expression value.

### R session information

```{r}
sessionInfo()
```
